---
output:
  bookdown::word_document2:
    reference_docx: "Styles.docx"
  
bibliography: references.bib
csl: springer-basic-author-date.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Contribution Title
 Mortality Modeling
 
# Authors
### Tim Riffe
 Research Scientist, Laboratory of Population Health, Max-Planck-Institute for Demographic Research, Rostock, Germany 
 
 Email: riffe@demogr.mpg.de

### Marília Nepomuceno
 Research Scientist, Research Group on Lifespan Inequalities, Max-Planck-Institute for Demographic Research, Rostock, Germany 
 
 Email: nepomuceno@demogr.mpg.de

### Ugofilippo Basellini
 PhD Researcher, Institut national d’études démographiques (INED), Paris, France and Interdisciplinary Centre on Population Dynamics (CPOP), Department of Public Health, University of Southern Denmark, Odense, Denmark 
 
 Email: ugofilippo.basellini@ined.fr  

# Synonyms
 Mortality Laws 

# Definition
Mortality models approximate mortality patterns or dynamics over age and or time. An age pattern of mortality can be any mathematical function of mortality, such as rates, probabilities, survivorship, or death distributions. Such functions may be modeled in the form of a life table or a simplified model of parameters. Mortality models in general fall into three main categories: (i) models designed to help understand regularities in mortality patterns and dynamics, for example where population-level mortality patterns are modeled as an emergent property of dynamics at the individual level, (ii) those that aim to predict mortality patterns, for example for purposes of pension forecasting, and (iii) those aimed at mortality measurement for purposes of mortality and health monitoring. In the following, mortality modeling refers to models of mortality measurement at the population level. 

# Overview
Mortality modeling has mainly grown in its ability to capture mortality changes over age and time. The first who explicitly modeled mortality was Graunt in [-@graunt1662natural] The author, through an empirical life table, analyzed mortality conditions in London. From Graunt [-@graunt1662natural] to date, models have become more sophisticated and complex.
 
Before the 20th century, mortality models were often developed for insurance reasons. This justify why models grown in its accuracy to measure adult mortality. In 1825, Gompertz documented one of the most well-known law of mortality. He theorized that adult mortality increases exponentially with age. Few decades later, Makeham (1867) proposed a modified version of the Gompertz model by including a parameter to capture the background mortality to better describe adult mortality. However, the unprecedented mortality improvements in the 20th century experienced by much of the world have motivated the development of mortality modeling.
 
The last century witnessed remarkable changes in patterns of health and disease. The leading causes of death shift from infectious and man-made to degenerative diseases [@omran1971theory]. A period dominated by changes in mortality among children and young adults moves towards to a period of changes in mortality at advanced ages. Overall, the focus of mortality modeling has been following this transition, and models that describe the entire age range has been developed. For instance, in 1980, Heligman and Pollard proposed a model by decomposing the age pattern of mortality into child, adult and old mortality. This model, in addition to cover the entire age range, it captures the accident hump in adulthood.
 
The sustained improvements in mortality at advanced ages in most industrialized countries increased the need for accurate measurement of old mortality.  Due to the very small number of survivors at the most advanced ages, and the scarce availability of accurate data for these individuals, modeled mortality at older ages is challenging. To date, logistic and exponential models has been used to describe the age pattern of mortality at very high ages. The first who proposed a logistic function to describe adult and old mortality was Perks [-@perks1932] Since then, Beard (-@beard1971), Kannisto (-@kannisto1994development), and Thatcher et al. (-@thatcher1998force) also theorized a slower rate of mortality increase at very old ages.
 
Over the centuries, theoretical models have often described the force of mortality (link to entry). From these models, one can derive several mortality functions such as the lifespan and the number of survivors at different ages. Thus, the ability of mortality models to reflect the dynamics of mortality has great implications in measures widely used for monitoring health. For example, the life expectancy at birth, that summarizes the risk of death over the entire age range, reflects the age-specific pattern of mortality implicit in the theoretical model. Higher life expectancy would derive from a logistic model (slower rate of mortality increase at advanced ages) compared to that derived from a model that assumes exponential mortality increase at older ages.
 
Mortality modeling is also important to forecast mortality. Most methods of mortality forecasting extrapolate both age patterns and trends over time. Extrapolative methods may also use summary mortality measures, such as life expectancy. Irrespective to the use of aggregate measures or the risk of death at each age, extrapolative methods make use of a mortality model, as described by a life table or a mathematical model. The Lee-Carter method (Lee Carter 1992) is a widely used extrapolative method to forecast mortality.
 
The evolution of mortality models moves towards to a better description of the age patterns of mortality. This process has the major advantage of producing accurate aggregate mortality measures, realistic figures of the mortality future, and better understanding of mechanisms behind the changes in the age pattern of mortality.

# Key Research Findings
This section contains an overview of some key findings that have been made in mortality modeling over the last centuries.

### Gompertz law 
One of the most important contributions to modeling mortality traces back to almost two centuries ago. The British mathematician and actuary Benjamin Gompertz [-@gompertz1825xxiv] theorized that the adult age-pattern of the force of mortality, specifically around the age range 30-90, increases exponentially with age. This assumption has proven to be a pretty accurate representation of the mortality pattern of several populations throughout the last two centuries. As such, the Gompertz law of mortality is today one of the most well-known models of human mortality, and it is still widely employed in demographic and actuarial analyses. The first panel of Fig \@ref(fig:Gompertz) shows the fit of the Gompertz model to the male population of Switzerland in 2010. As the graph shows, the assumption of exponential (linear in log scale) increase of mortality with age is still a valid representation of the human mortality pattern. 

```{r Gompertz, echo = FALSE, message = FALSE, warning = FALSE, fig.cap="Observed and fitted death rates using a log-linear (Gompertz) model for males aged 40 to 100 in Switzerland in the year 2010.", fig.width = 7, fig.height=7}
# This data object created in SSE.R
DX <- c(217, 17, 9, 8, 10, 7, 5, 4, 4, 11, 10, 6, 6, 8, 13, 12, 10, 
		25, 32, 37, 49, 45, 48, 33, 39, 48, 48, 58, 42, 54, 47, 59, 58, 
		64, 54, 66, 68, 74, 93, 74, 89, 88, 105, 112, 112, 105, 113, 
		135, 157, 165, 183, 185, 226, 262, 226, 259, 301, 333, 315, 342, 
		345, 352, 361, 404, 430, 467, 479, 530, 586, 636, 652, 690, 670, 
		750, 760, 866, 884, 950, 985, 1054, 1074, 976, 900, 920, 953, 
		983, 964, 898, 822, 712, 651, 613, 450, 395, 253, 195, 140, 87, 
		73, 39, 33, 17, 9, 5, 2, 0, 0, 0, 0, 0, 0)
EX <- c(39980.5, 40331, 40832.5, 41767.5, 42090.5, 41878.5, 42626.5, 
		43947, 45087.5, 45245.5, 44681, 44235.5, 43495, 42805.5, 42704, 
		42827, 42852.5, 42653.5, 42908, 42928, 42327, 41416.5, 41457, 
		41487.5, 41828, 43270.5, 44962, 46922, 49138.5, 51123, 53192, 
		55358, 57517.5, 59765.5, 62033.5, 63767, 63794, 62494, 60825, 
		59561.5, 58440.5, 57074.5, 56005.5, 55159.5, 53630, 51941, 50460.5, 
		49869.5, 49207.5, 49311.5, 49792.5, 49721.5, 49760.5, 49505.5, 
		48749.5, 47447.5, 45978.5, 44169, 41469, 38244, 36041, 34893, 
		33470.5, 32602.5, 32242, 31326.5, 29867, 28764, 27864, 27003.5, 
		25864.5, 24411.5, 23303.5, 22292.5, 21465.5, 20294.5, 19246.5, 
		18193, 17190.5, 16204, 13758.095, 11652.475, 10355.17, 9281.875, 
		8314.4, 7685.01, 6897.155, 5820.03, 4728.975, 3801.81, 3072.335, 
		2346.85, 1723.14, 1258.645, 871.085, 558.4, 357, 243, 152, 80.5, 
		53.5, 25.5, 15.5, 5.5, 1.5, 1.5, 1, 0.5, 0, 0, 0)
ages        <- 0:110
x           <- 40:100
mx.obs      <- DX / EX
x           <- 40:100
m           <- length(x)
Dx_sub      <- DX[ages%in%x]
Ex_sub      <- EX[ages%in%x]
mx.obs_sub  <- mx.obs[ages%in%x]

fit.gompertz <- MortalityLaws::MortalityLaw(x = x, Dx = Dx_sub, Ex = Ex_sub,
		law = "gompertz", opt.method = "poissonL")

# x             <- Gompertz$x
# mx.obs_sub    <- Gompertz$mx.obs_sub
# fitted.values <- Gompertz$fitted.values
my.cols.obs   <- c("grey20")
my.cols.mod   <- c("#d7191c")
par(mai=c(1,1,.1,.1))
plot(x,mx.obs_sub,log="y",type="n",axes=FALSE,xlab="",ylab="")
mtext("Age",side=1,line=2.5,cex=2)
mtext("Log-mortality",side=2,line=2.75,cex=1.75)
abline(h=c(0.002,0.01,0.05,0.5),lty=3,col="grey70")
abline(v=seq(40,100,10),lty=3,col="grey70")
axis(1);axis(2,las=2,at=c(0.002,0.01,0.05,0.5));box()
points(x,mx.obs_sub,pch=5,cex=1.5,col=my.cols.obs,lwd=2)
lines(x,fit.gompertz$fitted.values,col=my.cols.mod,lwd=2.5)
legend("bottomright",c("Observed","Gompertz"),col=c(my.cols.obs,my.cols.mod),
		lwd=2.5,bty="n",cex=1.5,pch = c(5,NA),lty=c(NA,1))
```

### Negative senescence and hump
Another key finding regarding the age-pattern of mortality was first described a few years after Gompertz. In 1871, the Danish astronomer and actuary Thorvald Thiele introduced the decomposition of the human mortality pattern into three groups that operate principally, or almost exclusively, upon childhood, middle, and old ages, respectively. These three components of mortality represent the risk of death at different times in life, and they have been observed to follow a stable pattern throughout history. The first component represents infant and child mortality, which decreases steadily with age, with a sharp reduction after birth. The second component, known as the accident hump for its particular shape, describes mortality after the onset of puberty and stretching into younger adult ages. Specifically, the component captures the accident deaths due to the risk-taking behavior of young men and, historically, mortality related to childbearing in women. Finally, the third component is generally denoted as senescent mortality, which was Gompertz’s centre of attention. As noted above, the component starts at middle-adult ages and increases exponentially with age. Several parametric and non-parametric approaches have been proposed in the last few decades to decompose the human mortality pattern. The second panel of Fig. \@ref(fig:SSE) shows this decomposition, applied to the same population as above. The three components of mortality clearly emerge from the graph. 


```{r SSE, echo = FALSE, message = FALSE, warning = FALSE, fig.cap="decomposition of the observed death rates into infant, young adult excess, and senescent components for males in Switzerland in the year 2010.", fig.width = 7, fig.height=7}
# This data object created in SSE.R
library(MortalitySmooth)
source("R/SSE.R")
ages   <- 0:110
years  <- 2000:2001
cou    <- "Switzerland"
DX     <- selectHMDdata(country = cou,data = "Deaths",sex = "Males",
                     ages = ages, years = years)
EX     <- selectHMDdata(country = cou,data = "Exposures",sex = "Males",
                     ages = ages, years = years)
fitSSE <- SSEdecomposition(Z = DX, E = EX, ages = ages, years = years)
jj          <- 1
mx.obs      <- DX[,jj] / EX[,jj]
my.cols.obs <- c("grey40")
#par(mar=c(3.75,4.25,1,1))
## bottom, left, top, right
plot(ages,mx.obs,log="y",t="n",axes="F",xlab="",ylab="",ylim = range(exp(-11),mx.obs[!mx.obs==0],na.rm=T))
mtext("Age",side=1,line=2.5,cex=2)
mtext("Log-mortality",side=2,line=2.75,cex=1.75)
abline(h=c(1e-4,0.001,0.01,0.1,1),lty=3,col="grey70")
abline(v=seq(0,100,20),lty=3,col="grey70")
axis(1);
axis(2,las=2,at=c(1e-4,0.001,0.01,0.1,1),
     labels=c(1e-4,0.001,0.01,0.1,1));box()
points(ages,mx.obs,pch=5,cex=1.25,col=my.cols.obs,lwd=2)
lines(ages,exp(fitSSE$LHAZhat[,jj]),col=1,lwd=3)
lines(fitSSE$xINF,exp(fitSSE$LHAZ1hat[,jj]),col=6,lwd=2.5,lty=2)
lines(fitSSE$xSEN,exp(fitSSE$LHAZ2hat[,jj]),col=3,lwd=2.5,lty=2)
lines(fitSSE$xACC,exp(fitSSE$LHAZ3hat[,jj]),col=4,lwd=2.5,lty=2)
legend("bottomright",c("Observed","Infant","Accident","Senescent"),col=c(my.cols.obs,6,4,3),
       lwd=2.5,bty="n",cex=1.5,pch = c(5,NA,NA,NA),lty=c(NA,2,2,2))

```

### Debate on plateau 
There is a general agreement that mortality increases exponentially from mid-adult to ages 80-90, as described by the Gompertz law. However, there is not yet a consensus regarding the mortality trajectory at the most advanced ages. Some studies suggested that the exponential growth of mortality with age is followed by a period of deceleration, with slower rates of mortality increase at the oldest ages, creating a plateau of human mortality (@barbi2018plateau, Gampe 2010; Robine and Vaupel 2001; Robine et al. 2005). Another group of researchers claimed that the mortality deceleration in later life or the mortality plateau is more expressed for data with lower quality, and hence mortality continues to grow exponentially at the highest ages (Gavirilov and Gavirilova 2011, 2015, 2019). Logistic curves like Kannisto (Kannisto 1994), Beard (1971) and Perks (Perks 1932) model the mortality plateau, while the Gompertz law rises exponentially with age.

### Rectangularization 
Rectangularization is defined by the process when the shape of the survival curve becomes more rectangular due to reductions in premature mortality and the concentration of deaths at older ages. In 1980, James Fries claimed that the rectangularization occurs when life expectancy at birth approaches the average lifespan due to a decrease in variability in age-at-death (Fries 1980). In other words, when lifespan variability decreases, deaths are compressed at older ages. However, since the second half of the twentieth century, lifespan variability has been stagnating and life expectancy continued to increase in high-income countries, resulting in a process known by mortality shifting (Kannisto 1996; Boongarts 2005; Bergeron-Boucher et al. 2015). Mortality shifting occurs when life expectancy increases due to a shift in the death distribution toward older ages with nearly constant lifespan variability (Vaupel 1986, Yashin et al. 2001, Bongaarts 2005, Canudas-Romo 2008). 

### Oeppen-Vaupel line 
Oeppen and Vaupel [-@oeppen2002demography] showed one of the most striking regularities observed in human mortality during the last centuries: the highest level of observed life expectancy has been rising at a steady pace of almost 3 months per year during the last 160 years. The authors analyzed the evolution of the female best-practice life expectancy at birth (that is, the life expectancy of the country holding the highest level in the whole world in a given calendar year) from 1840 to 2000, finding that the trend is very linear, rising at a steady pace of almost 3 months per year during the last 160 years. Fig. \@ref(fig:OV) shows this remarkable finding, which has had direct implications in a number of spheres, for example, mortality forecasting. The best-practice line has indeed been employed in some forecasting methodologies (cite).   

```{r OV, echo = FALSE, message = FALSE, warning = FALSE, fig.cap="Best-practice female life expectancy at birth from 1840 to 2016. data: HMD.", fig.width = 7, fig.height=7 }
library(ggplot2)
my.cols <- c("purple3",
  "dodgerblue4", "green4","orange2", "firebrick3","blue4")

source("Data/OV.R")
ggplot(data=df,aes(x=year,y=e0max)) + 
  geom_point(aes(shape=Country, color=Country),size=3.5) + 
  scale_shape_manual(values=c(2, 5, 1, 0, 8, 6)) +
  scale_color_manual(values=my.cols) +
  xlab("Year") + ylab("Life expectancy") + 
  geom_smooth(method = "lm", se = FALSE, color= "black") +
  theme(legend.title = element_text(size=18,face="bold"),
        axis.title.x = element_text(size=16,face="bold"),
        axis.title.y = element_text(size=16,face="bold"),
        axis.text.x = element_text(size=14),
        axis.text.y = element_text(size=14)) 
```

# Examples of Application 	 	 	
### Model lifetables 
Model lifetables comprise a family of techniques designed to relate partial mortality information to a full age pattern of mortality, which can then be used to calculate standard mortality indicators such as life expectancy. Many variants exist and have long been in use, especially in the context of mortality measurement in countries without complete vital registration [@unpd1982; @wpp2012; @coale1983regional; @wilmoth2012flexible; @clark2015singular]. Model patterns are derived by reducing the mortality patterns in a large collection of lifetables into a smaller set of relationships that when scaled and warped according to a small number of (usually) intuitive parameters.

```{r MLTwilmoth, echo = FALSE, message = FALSE, warning = FALSE, fig.cap="A demonstration of some of the flexibility given in the Wilmoth et al (2012) model lifetable approach. A standard is derived from HMD male lifetables. From this, nine different survivorships are derived from combinations of specified child and adult mortality. Three levels of child mortality (0.01,0.05,0.10) are paired with three levels of relative adult mortality, 1.5, 2.5, and 5.0 times higher than the given child mortality estimate, respectively.", fig.width = 7, fig.height=7}
# Note, full code to derive the object LX is located
# 
library(RColorBrewer)
# lower age bounds
x <- c(0,1, seq(5, 110, by = 5))
source("Data/LX.R")

# some colors to vary
reds    <- brewer.pal(5,"Reds")[3:5]
blues   <- brewer.pal(5,"Blues")[3:5]
purples <- brewer.pal(5,"Purples")[3:5]
CC      <- rbind(blues,purples,reds)
matplot(x, LX, lwd = 2, col = t(CC), lty = 1:3,type='l',
		xlab="Age",ylab="Survival probability", las = 1)
text(3,LX[3, c(1,4,7)]+.015, c("0.01","0.05","0.10"), pos = 4)
text(0,.7,"child mortality\n(ages 0-5)",pos=4)
arrows(4,.76,8,.88, length = .08)

text(28,LX[8, c(7:9)]+ c(.01,-.03,-.06), c("1.5x","2.5x","5.0x") , pos = 4)
text(20,.5,"adult mortality\n(ages 15-65)\nrelative to child",pos=4)
arrows(30,.6,32,.7, length = .08)
```

Fig. \@ref(fig:MLTwilmoth) demonstrates the flexibility that most model systems can achieve, in this case using the Wilmoth et al approach [-@wilmoth2012flexible], where the standard mortality pattern is derived from male lifetables in the Human Mortality Database [-@HMD]. Results are calculated using the MortalityEstimate R package [@MortalityEstimate2019]. In practice the standard could be derived from any corpus of lifetables. The standard is then adjusted according to specified estimates of child and adult mortality. Child mortality is defined as the probability of not reaching age 5, and adult mortality is here defined as the probability of a 15-year-old not reaching age 65. These are just two of several potential parameterizations of this model. These combinations reveal a wide range of realistic mortality patterns that can be derived from just two parameters that can be estimated independently.

For the case of gerontological research, model lifetables such as that demonstrated can 1) extend the geographical range of research to populations not usually included in research due to data limitations, 2) extend estimation to subpopulations for which complete lifetables cannot be directly calculated, 3) systematize the relationship between mortality in different age groups to infer older age mortality from younger ages when older ages are not directly observed (compare with extrapolation using parametric mortality laws).

### Extrapolation and smoothing 
The Human Mortality Database (HMD) uses a parametric procedure to smooth old-age death rates in period life tables. Specifically, the Kannisto model of old-age mortality [@thatcher1998force] is used to remove the randomness inherent to the observed rates at older ages, thus obtaining “an improved representation of the underlying mortality conditions” [@hmd2017methods, p. 34]. The model is fitted to ages 80 and above, separately for males and females, assuming a Poisson distribution for death counts. The estimated parameters yields smooth death rates, which are used to replace the observed rates for all ages above age . In particular,  is defined as the lowest age where there are at most 100 deaths (male or female), but it is constrained to fall within the age interval [80, 95] (for additional information, see @hmd2017methods).

```{r Extrap, echo = FALSE, message = FALSE, warning = FALSE, fig.cap="Observed and lifetable death rates for females in England and Wales in the year 2010, with fit of four different parametric mortality models. The Gompertz model is log linear, whereas the Kannisto and Gamma-Gompertz models deviate after age 95, tapering toward a plateau. The Weibull extrapolation is intermediate in this case.", fig.width = 7, fig.height=7}
library(MortalityLaws)
# load GBRENW 2010 female data
source("Data/HMD.R")
ages 			 <- 0:110
year             <- 2010
cou              <- "GBRTENW"
names(Dx)        <- names(Ex) <- ages
mx.obs           <- Dx / Ex
## focus on old age mortality
x.start          <- 80
x                <- x.start:tail(ages,1)
m                <- length(x)
Dx_sub           <- Dx[ages%in%x]
Ex_sub           <- Ex[ages%in%x]
mx.obs_sub       <- mx.obs[ages%in%x]
mx.lt_sub        <- mx.lt[ages%in%x]

## select four models to fit to the data
models <- c("gompertz","ggompertz","kannisto","weibull")
fit.gompertz <- MortalityLaw(x = x, Dx = Dx_sub, Ex = Ex_sub,
                             law = models[1], opt.method = "poissonL")
fit.ggompertz <- MortalityLaw(x = x, Dx = Dx_sub, Ex = Ex_sub,
                              law = models[2], opt.method = "poissonL")
fit.kannisto <- MortalityLaw(x = x, Dx = Dx_sub, Ex = Ex_sub,
                             law = models[3], opt.method = "poissonL")
fit.weibull <- MortalityLaw(x = x, Dx = Dx_sub, Ex = Ex_sub,
                            law = models[4], opt.method = "poissonL")

## choose colors
my.cols.obs <- c("grey40","grey20")
#my.cols.mod <- c("#d7191c","chartreuse4","darkorange","dodgerblue3")
my.cols.mod <- c("#008E77","#458200","#A46400","#C23F86")
my.lty.mod  <- c("81","44","63","8222")

plot(x,mx.obs_sub,log="y",t="n",axes="F",xlab="",ylab="")
mtext("Age",side=1,line=2.5,cex=2)
mtext("Log-mortality",side=2,line=2.75,cex=1.75)
abline(h=c(0.05,0.1,0.2,0.5,1),lty=3,col="grey70")
abline(v=seq(80,110,5),lty=3,col="grey70")
axis(1);axis(2,las=2);box()
points(x,mx.obs_sub,pch=5,cex=1.5,col=my.cols.obs[2],lwd=2)
points(x,mx.lt_sub,pch=1,cex=1.7,col=my.cols.obs[1],lwd=2)
legend("topleft",c("Observed","Life Table"),col=rev(my.cols.obs),
       pch=c(5,1),bty="n",cex=1.5,lwd=2,lty=NA,pt.cex=c(2,2.25))
matlines(x,cbind(fit.gompertz$fitted.values,
				 fit.weibull$fitted.values,
				 fit.ggompertz$fitted.values,
				 fit.kannisto$fitted.values),
		 col=my.cols.mod,
		 lwd=2.5,
		 lty=my.lty.mod)
legend("bottomright",c("Gompertz","Weibull","Gamma-Gomp","Kannisto"),col=my.cols.mod,
       lwd=2.5,lty=my.lty.mod,bty="n",cex=1.5)
```

The right panel of Fig. \@ref(fig:Extrap) shows an example of this procedure. This figure shows the observed and lifetable death rates for females in England and Wales in 2010 (HMD). The two rates depart from each other after age 95 (which is age for this population, as less than 100 deaths were observed from the age-group 106). In particular, the life table death rates are derived from the Kannisto mortality model, as discussed above. In addition, three other well-known models of mortality to this population: the Gompertz, the Weibull and the Gamma-Gompertz model. The fitted lines show that the Gamma-Gompertz model produces estimates that are very close to the Kannisto model, while the Gompertz and Weibull model overestimate the observed pattern of mortality at the highest ages. 

# Future Directions of Research
Future research in mortality modeling will extend the applications of presently available modeling techniques to yet-understudied populations, either due to innovations to increase applicability or due to new data becoming available. Several thematic areas for future research are evident:

### Limits to life and the mortality plateau 
The scientific community is far from reaching a consensus on whether and in what way the human longevity is subject to fundamental limits, either in terms of the maximum age attained or a maximum death rate (compare @olshansky1990search and @barbi2018plateau) , and this is a fundamental concern for the field of gerontology. For example, approaches based on the consideration of physiological constraints have not been harmonized with demographic approaches. Data collection and validation of centenarians and super-centenarians is ongoing, and the number of people reaching these ages worldwide has been increasing rapidly [@vaupel2010biodemography]. In the coming decades the increased amount of data on the longest lived will be able to provide more nuance to the question of whether or not and at what level mortality levels off in old age, and ideally lead to a consensus between disciplines on the character of mortality among the extremely longevous.

###  New population definitions 
Many methods to harmonize mortality estimation from diverse sources and to make deficient or incomplete vital register data more usable in standard lifetable applications. The same push to improve mortality estimation worldwide is also relevant to populations with good data, which can benefit from the same tools in order to differentiate mortality outcomes for small, highly local, difficult-to-observe, or partially observed subpopulations. Large populations may be further stratified in creative ways (for example, by different definitions of income, wealth, and capital) that might reveal group differences in mortality patterns and levels. Likewise, new registers will better identify health risk factors that differentiate mortality outcomes.

###  Coherent modeling 
Innovation is needed to overcome modeling difficulties arising from the two-sided compositional nature of mortality: First, mortality is measured with respect to some definition of population at risk, which itself is necessarily a composition of heterogeneous risk levels. A standard approach to model the unobserved risk structure of mortality would benefit the field greatly. Second, mortality is a composition of outcomes, such as causes of death. This raises difficulties in practice, for example in projecting mortality by cause of death, or when smoothing mortality jointly by cause of death. Such operations would also benefit from standard solutions.

###  Lifespan inequality 
Most mortality research has been focused on life expectancy, or average length of life as the primary lifetable outcome, but recent research highlights the necessity of estimating and monitoring lifespan uncertainty as a primary indicator of population health status, and as a fundamental kind of population inequality [@van2018case]. Mortality modeling often relates parameters to life expectancy, but not lifespan inequality, and lifespan inequality is rarely used to calibrate models [@bohk2017lifespan]. Future research will measure social differences in lifespan inequality, and modeling will either account for inequality or be parameterized in terms of inequality.

###  Model translation 
To the extent that mortality modeling has bifurcated into subfields for the purposes of measurement, prediction, and explanation, there is potential for innovation by translating models to other ends. For example, Sharrow and Anderson [-@sharrow2016quantifying] offer a process-based model to partition mortality into intrinsic and extrinsic components, and Camarda et al [-@camarda2016sums] propose a decomposition of mortality into three substantive components for infant, young adult excess, and old-age mortality. Both are in the first place a fundamental question of measurement, but hold the potential to offer new approaches to prediction or relate to models of understanding. Many instances of model cross-fertilization of this kind are likely to be proposed in future research.

# Summary

# Cross references

# References
